---
title: "Cluster-mass for EEG data"
author: "Angela Andreella"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clustermass_EEG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  echo = TRUE,
  warnings = FALSE,
  out.width = "500px",
  dpi=150
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

We will explain and apply in ``R`` the **permutation-based cluster-Mass** method proposed by [Maris and Oostenveld, 2007](https://doi.org/10.1016/j.jneumeth.2007.03.024) and developed in the ``R`` package ``permuco4brain`` by [Frossard and Renaud, 2018](https://cran.r-project.org/web/packages/permuco/vignettes/permuco_tutorial.pdf), using electroencephalography (EEG) data. The cluster-mass is computed considering: 

- the time series of one channel: **temporal cluster-Mass**; 
- the time series of multiple channels: **spatial-temporal cluster-mass**. 

Finally, the **All-Resolution Inference** (ARI) from [Rosenblatt et al. 2018](https://doi.org/10.1016/j.neuroimage.2018.07.060) is applied to compute the lower bound for the true discovery proportion inside the clusters computed. Here, we will use the ``ARIeeg`` and ``hommel`` ``R`` packages.

## Packages

First of all, you need to install and load the following packages:

```{r 1, warning=FALSE,message=FALSE}
#devtools::install_github("angeella/ARIeeg")
#devtools::install_github("bnicenboim/eeguana")
#devtools::install_github("jaromilfrossard/permuco4brain")
#devtools::install_github("jaromilfrossard/permuco")
#devtools::install_github("livioivl/eegusta")
library(ARIeeg) #to compute ARI for spatial-temporal clusters
library(eeguana) #to manage eeg data
library(eegusta) #to manage eeg data
library(permuco4brain) #to compute the spatial-temporal clusters
library(dplyr) 
library(ggplot2)
library(plotly)
library(tidyverse)
library(permuco) #to compute the temporal clusters
library(hommel) #to compute ARI for temporal clusters
library(abind)
```

## Data

The Dataset from the package ```ARIeeg``` is an **ERP experiment** composed by:

- 20 Subjects;
- 32 Channels;
- Stimuli: pictures. Conditions:
    1. (f): fear (face);
    2. (h): happiness (face);
    3. (d): disgust (face);
    4. (n): neutral (face);
    5. (o): object.

You can load it using:

```{r 2}
load(system.file("extdata", "data_eeg_emotion.RData", package = "ARIeeg"))
```

We transform the data as ```eeg_lst``` class object from the ``R`` package ```eeguana``` using the function ``eegUtils2eeguana`` from the ``R`` package ``eegusta``:

```{r 3}
dati_lst = eegUtils2eeguana(data = dati)
is_eeg_lst(dati_lst) #check
```

and we drop off the final five channels: 

```{r 4}
chan_to_rm <- c("RM"  ,  "EOGvo" ,"EOGvu"
                , "EOGhl", "EOGhr")
dati_lst <- 
  dati_lst %>%
  dplyr::select(-one_of(chan_to_rm))
```

Finally, we segment the data and select two conditions, i.e., **disgust face**(number $3$) and **object** (number $5$):

```{r 5, warning=FALSE,message=FALSE}
data_seg <- dati_lst %>%
  eeg_segment(.description %in% c(3,5),
              .lim = c(min(dati$timings$time), max(dati$timings$time))
  ) %>% eeg_baseline()  %>%
  mutate(
    condition =
      description
  ) %>%
  dplyr::select(-c(type,description))
```

Some plots to understand the global mean difference between the two conditions:

```{r}
A<-data_seg %>%
  select(Fp1,Fp2, F3, F4) %>%
  ggplot(aes(x = .time, y = .value)) +
  geom_line(aes(group = condition))  +
  stat_summary(
    fun = "mean", geom = "line", alpha = 1, size = 1,
    aes(color = condition),show.legend = TRUE
  ) +
  facet_wrap(~.key) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = .17, linetype = "dotted") +
  theme(legend.position = "bottom")+
  scale_color_manual(labels = c("Disgust", "Object"), values = c("#80bfff", "#ff8080"))
A
```

if you want an interactive plot, you can use the function ``ggplotly`` from the package ``plotly``:

```{r, warning=FALSE}
plotly::ggplotly(A)    
```

## Theory

### Multiple testing problem?

The aim is to test if the difference in brain signal during the two conditions is different from $0$ for each time point, i.e., $500$. If the full set of channels is considered, we also have tests for each channel, i.e., $27$, returning a total number of tests equals $500 \cdot 27$. Therefore, we have $500$ or $500 \cdot 27$ statistical tests to perform at group-level, so considering the **random subject effect**. The multiple testing problem is obvious, and correction methods such as Bonferroni or similar do not capture the time(-spatial) correlation structure of the statistical tests; it will be a conservative method.

The cluster mass method is then used, proposed by [Maris and Oostenveld, 2007](https://doi.org/10.1016/j.jneumeth.2007.03.024). It is based on **permutation theory**;  it gains some power with respect to other procedures correcting at the (spatial-)temporal cluster level instead of at the level of single tests. It is similar to the cluster mass approach in the fMRI framework, but in this case, the *voxels*, i.e., the single object of the analysis, are expressed in terms of time-points or combination time-points/channels. The method can then gain some power with respect to some traditional conservative FWER correction methods exploiting the (spatial-)temporal structure of the data.

### Repeated Measures ANOVA Model

The cluster mass method is based on the **repeated measures ANOVA** model, i.e.,

$$
y = 1_{N \times 1} \mu +  X^{\eta} \eta  +  X^{\pi}\pi +   X^{\eta \pi}\eta \pi + \epsilon
$$

where $1_{N \times 1}$ is a matrix with ones and

  1. $\mu$ is the **intercept**;
  2. $y \in \mathbb{R}^{N \times 1}$ is the response variables, i.e., the **signal**, in our case $N = n_{subj} \times n_{stimuli} = 40$;
  3. $X^{\eta} \in \mathbb{R}^{N \times n_{stimuli}}$ is the **design matrix** describing the **fixed effect** regarding the stimuli, and $\eta \in \mathbb{R}^{n_{stimuli} \times 1}$ the corresponding parameter of interest;
  4. $X^{\pi} \in \mathbb{R}^{N \times n_{subj}}$ is the **design matrix** describing the **random effect** regarding the subjects, and $\pi \in \mathbb{R}^{n_{subj} \times 1}$ the corresponding parameter.
  5. $X^{\eta \pi}$ is the **design matrix** describing the **interaction effects** between subjects and conditions, and $\eta \pi$ the corresponding parameter;
  6. $\epsilon \in \mathbb{R}^{N \times 1}$ is the **error term** with $0$ mean and variance $\sigma^2 I_N$.

Therefore, $y \sim (1\mu + X^{\eta} \eta, \Sigma)$, $\pi \sim (0, \sigma^2_{\pi} I_{n_{subj}})$ and $\eta \pi \sim (0,\text{cov}(\eta \pi))$. (N.B: $\eta \pi$ is not the product between $\eta$ and $\pi$ but refers to the **interaction effects** between subjects and conditions).

We want to make inference on $\eta$, such that $H_0: \eta = 0$ vs $H_1: \eta \ne 0$. We do that using the **F statistic**, i.e.,

$$
F = \dfrac{y^\top H_{X^{\eta}} y / (n_{stimuli} - 1)}{ y^\top H_{X^{\eta \pi}}y/(n_{stimuli} -1)(n_{subj} -1)} 
$$
where $H_{X}$ is the **projection matrix**, i.e., $H_{X} = X(X^\top X)^{-1} X^\top$. In order to compute this test, we use an alternative definition of $F$ based on the residuals:

$$
F_r = \dfrac{r^\top H_{X^{\eta}} r / (n_{stimuli} - 1)}{ r^\top H_{X^{\eta \pi}}r/(n_{stimuli} -1)(n_{subj} -1)} 
$$

where $r = (H_{X^{\eta}} + H_{X^{\eta\pi}})y$. For further details, see [Kherad Pajouh and Renaud, 2014](https://link.springer.com/article/10.1007/s00362-014-0617-3).

So, let the group of permutation, including the identity transformation, $\mathcal{P}$, we use $r^\star = P r$, where $P \in \mathcal{P}$ to compute the null distribution of our test, i.e., $\mathcal{R}$, and then the p-value, i.e.,

$$
\text{p-value} = \dfrac{1}{B} \sum_{F^\star_r \in \mathcal{R}} \mathbb{I}(|F^\star_r| \ge |F_r|)
$$

if the two-tailed is considered, where $F^\star_r = f(r^\star)$.

We have this model for each time point $t \in \{1, \dots, 500\}$ and each channel, so finally we will have $n_{\text{time-points}} \times n_{\text{channels}}$ statistical tests/p-values (raw).

### Temporal Cluster mass

This method has been proposed by [Maris and Oostenveld, 2007](https://doi.org/10.1016/j.jneumeth.2007.03.024). It assumes that an effect will appear in clusters of adjacent time frames. Having statistics for each time point, we form these clusters using a threshold $\tau$ as follows:

<div style="text-align:center" markdown="1">
![Example of cluster mass EEG from [Frossard, 2019](10.13097/archive-ouverte/unige:125617)](Image/clusterMass.JPG){width=50%}
</div>

All contiguous time points with statistics above this threshold define a single cluster $C_i$ with $i \in \{1, \dots, n_C\}$, where $n_C$ is the number of clusters found. For each time point in the same cluster $C_i$, we assign the same cluster mass statistic $m_i = f(C_i)$, where $f$ is the function that summarizes the statistics of the entire cluster. Typically, it is the sum of the $F$ statistics. The null distribution of the cluster mass $\mathcal{M}$ is computed by iterating the above process for each permutation. The contribution of a permutation to the cluster-mass null distribution is the maximum overall cluster mass of that permutation. To check the significance of the cluster $C_i$ of interest, we compare its cluster mass $m_i = f(C_i)$ with the cluster mass null distribution $\mathcal{M}$. Therefore, for each cluster $C_i$, we have the associated p-values computed as

$$
p_i = \dfrac{1}{n_P} \sum_{m^\star \in \mathcal{M}} I\{m^\star  \ge m_i\}
$$

where $m^\star \in \mathcal{M}$ is then calculated given permutation statistics. This method makes sense when analysing EEG data because if a difference in brain activity is thought to occur at time $s$ for a given factor, then it is very likely that this difference will also occur at time $s + 1$ (or $s - 1$).

### Spatial-temporal Cluster mass 

In this case, we use graph theory, where the vertices represent the channels and the edges represent the **adjacency relationships** between two channels. The adjacency must be defined using prior information, so the three-dimensional Euclidean distance between channels is used. Two channels are defined as adjacent if their Euclidean distance is less than the threshold $\delta$, where $\delta$ is the smallest Euclidean distance that yields a connected graph [Cheval, et al., ](https://www.sciencedirect.com/science/article/pii/S0028393218303981?casa_token=WZEwrKRyNWkAAAAA:tYPdd5QarUoc8ErySnlzcPgr7c3naUjct62Uv5Cf3Dh4t6RsszLN26hIncUSZGHV82pInzcFSQ)). This follows from the fact that there is no unconnected subgraph for a connected graph. The existence of subgraphs implies that some tests cannot be enter in the same cluster, which is not a useful assumption for the present analysis ([Frossard and Renaud, 2018](https://cran.r-project.org/web/packages/permuco/vignettes/permuco_tutorial.pdf); [Frossard, 2019](10.13097/archive-ouverte/unige:125617)).

Once we have a definition of spatial contiguity, we need to define temporal contiguity. We reproduce this graph $n_{\text{time-points}}$ times, and we have edges between pairs of two vertices associated to the same electrode if they are temporally adjacent. The final graph has a total number of vertices, i.e., number of tests, equals ($n_{\text{channels}} \times n_{\text{time-points}}$). The following figure shows an example with $64$ channels and $3$ time measures:

<div style="text-align:center" markdown="1">
![Example of graph of adjacency from [Frossard, 2019](10.13097/archive-ouverte/unige:125617)](Image/cluster.JPG){width=50%}
</div>

We then delete all the vertices in which statistics are below a threshold, e.g., the $95$ percentile of the null distribution of the $F$ statistics. So, we have a new graph composed of **multiple connected components**, where each connected component defines the spatial-temporal cluster. We compute for each spatial-temporal cluster the cluster-mass statistic as before.

The cluster-mass null distribution is calculated using permutations that preserve the spatial-temporal correlation structure of the statistical tests, i.e., no changing the position of the electrodes and mixing the time points.
We construct a three-dimensional array, where the first dimension represents the design of our experiments (subjects of $\times$ stimuli), the second one the time points, and the third one the electrodes. So, we apply permutations only in the first dimension using the method proposed by [Kherad Pajouh and Renaud, 2014](https://link.springer.com/article/10.1007/s00362-014-0617-3). 

## Application

In R, all of this is possible thanks to the ``permuco`` and ``permuco4brain`` packages developed by [Frossard and Renaud, 2018](https://cran.r-project.org/web/packages/permuco/vignettes/permuco_tutorial.pdf).

### Temporal Cluster-Mass

So, we select one channel from our dataset, e.g. the **Fp1**:

```{r}
Fp1 <- data_seg %>% select(Fp1)
```

1. Construct the $y$. We need to construct the three-dimensional **signal matrix**, having dimensions $40 \times 500$:

```{r}
signal_Fp1 <- Fp1%>%
    signal_tbl()%>%
    group_by(.id)%>%
    nest()%>% #creates a list-column of 40 data frames having dim 500 x 2
    mutate(data = map(data,~as.matrix(.x[-1])))%>% #drop off the first column of each dataframe, i.e., we take the column oof the signal for channel Fp1
    pull(data)%>% #takes data created in mutate
    invoke(abind,.,along = 2)%>% #we merge each dataframe in order to have one db having dimension 500 x 40
    aperm(c(2,1)) #traspose
dim(signal_Fp1) 
```

So, ``signal_Fp1`` is a data frame that expresses the channel **Fp1** signals under two conditions in $500$ time points for $20$ subjects.

2. Construct the $X^{\eta \pi}$, having dimensions $40 \times 2$:

```{r}
design <- 
  segments_tbl(Fp1)%>%
  select(participant_id, condition)
dim(design)
```

3. Define the **repeated measures ANOVA formula**: 

```{r}
f <- signal_Fp1 ~ condition + Error(participant_id/(condition))
```

In the formula, we need to specify the ``Error(.)`` term since we are dealing with a repeated measures design. We specify a subject-level random effect and a condition fixed effect nested within subjects.

Thanks to the ``permuco`` package, we can apply the temporal cluster-Mass for the channel **Fp1**:

```{r}

lm_Fp1 <- clusterlm(f,data = design)
summary(lm_Fp1)
```

Here we can see:

- The threshold used to construct the temporal clusters, i.e., `r lm_Fp1$threshold`;
- The type of cluster mass function, i.e., the sum of single statistical tests time contiguous; 
- When the cluster starts and ends, the value of the cluster mass and the associated corrected p-values.

For example, considering the first significant cluster, we can compute the cluster mass as:

```{r}
sum(lm_Fp1$multiple_comparison$condition$uncorrected$main[c(170:210), "statistic"])
```


We can also plot the temporal clusters:

```{r}
plot(lm_Fp1)
```

The red dots represent the significant temporal cluster for the channel **Fp1** composed by the time points from $170$ to $210$ using a threshold equals $4.098$. 

#### ARI in EEG cluster mass

However, our significant cluster says only that at least one test is different from $0$, we don't know how many tests/time-points are significant (**spatial specificity paradox**). So, we can apply ARI to understand the lower bound of the number of true discovery proportion. The cluster is composed by the time points from $170$ to $210$, i.e., the size of the cluster is equal to $41$.

```{r}
praw <- lm_Fp1$multiple_comparison$condition$uncorrected$main[,2]
cluster <- c(170:210)

discoveries(hommel(praw), ix = cluster)/length(cluster)*100
```

Therefore, we have at least `r round(discoveries(hommel(praw), ix = cluster)/length(cluster)*100,3)`$\%$ of true active time points in the cluster computed.

### Spatial-Temporal Cluster-Mass

1. Construct the $y$. We need to construct the three-dimensional **signal array**, having dimensions $40 \times 500 \times 27$:

```{r 7}
signal <- 
    data_seg%>%
    signal_tbl()%>%
    group_by(.id)%>%
    nest()%>%
    mutate(data = map(data,~as.matrix(.x[-1])))%>%
    pull(data)%>%
    invoke(abind,.,along = 3)%>%
    aperm(c(3,1,2))

dim(signal)
```

2. Construct the $X^{\eta \pi}$, having dimensions $40 \times 2$:

```{r 8}
design <- 
  segments_tbl(data_seg)%>%
  select(participant_id, condition)
dim(design)
```

3. Construct the **graph**, using $\delta = 53mm$ (the maximal distance for adjacency of two channels) and the function ``position_to_graph`` from the ``permuco4brain`` package:

```{r fig.align="center"}
graph <- position_to_graph(channels_tbl(data_seg), name = .channel, delta = 53,
                             x = .x, y = .y, z = .z)
plot(graph)
```

4. Define the **repeated measures ANOVA formula**: 

```{r 10}
f <- signal ~ condition + Error(participant_id/(condition))
```

Finally, run the main function:

```{r}
model <- permuco4brain::brainperm(formula = f,
                                  data = design,
                                  graph = graph,
                                  np = 1000,
                                  multcomp = "clustermass",
                                  return_distribution = TRUE)
```

where np indicates the number of permutation.

Then, we can analyze the output:

```{r}
print(model)
```

We have only two significant clusters. The first one is composed by $25$ channels while the second one by $8$ channels, with main channels P7. You can see in details the components of this cluster in

```{r}
head(names(model$multiple_comparison$condition$clustermass$cluster$membership[which(as.vector(model$multiple_comparison$condition$clustermass$cluster$membership)==3)]))
```

You can see the significant cluster (in red) at fixed time points (e.g. 300) using plot:

```{r}
plot(model, samples = 300)
```

and the significant cluster over time and over channels using:

```{r}
image(model)
```

where the significant clusters are represented in a colour-scale and the non-significant one in grey. The white pixels are tests which statistic are below the threshold.

#### ARI in EEG cluster mass

However, our significant clusters say only that at least one combination channels/time-points is different from $0$, we do not know how many combinations are significant (**spatial specificity paradox**). So, we can apply ARI to understand the lower bound of the number of true discovery proportion:

```{r}
ARIeeg::ARIeeg(model = model)
```


# References

 - Maris, E., & Oostenveld, R. (2007). Nonparametric statistical testing of EEG-and MEG-data. Journal of neuroscience methods, 164(1), 177-190.

 - Kherad-Pajouh, S., & Renaud, O. (2015). A general permutation approach for analyzing repeated measures ANOVA and mixed-model designs. Statistical Papers, 56(4), 947-967.
 
 - Frossard, J. (2019). Permutation tests and multiple comparisons in the linear models and mixed linear models, with extension to experiments using electroencephalography. DOI: 10.13097/archive-ouverte/unige:125617.
 
 - Frossard, J. & O. Renaud (2018). Permuco: Permutation Tests for Regression, (Repeated Measures) ANOVA/ANCOVA and Comparison of Signals. R Packages.
 
- Cheval, Boris, et al. "Avoiding sedentary behaviors requires more cortical resources than avoiding physical activity: An EEG study." Neuropsychologia 119 (2018): 68-80.
